---
title: "From Python to Base R: Institutional & Field-of-Study Analytics"
author: "Will, Leo, James"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    df_print: default
---

## Goals

-   Translate Python to base R, no tidyverse or dplyr.
-   Keep the **same analysis intent**: cleaning, profiling, mapping codes, ROI, regions, locale, and field-of-study.
-   Provide **side-by-side explanations**

------------------------------------------------------------------------

## Conventions Used

-   Use `read.csv()` and `read.delim()` to import data.
-   Use **named vectors** for code maps.
-   Use `as.numeric()`, `is.na()`, `complete.cases()` for cleaning.
-   Use `aggregate()`, `tapply()`, `by()` for grouped summaries.
-   Use **base graphics**: `plot()`, `points()`, `barplot()`, `boxplot()`, `legend()`.

------------------------------------------------------------------------

## 1) Institutional Columns to Load

```{r setup-paths-and-columns}
data_dir <- "College_Scorecard_Raw_Data_05192025"

columns_to_load_institution <- c(
  "UNITID","INSTNM","CONTROL","PREDDEG","HIGHDEG",
  "REGION","LOCALE","UGDS",
  "COSTT4_A","COSTT4_P",
  "TUITIONFEE_IN","TUITIONFEE_OUT","TUITIONFEE_PROG",
  "MD_EARN_WNE_P10",
  "NPT4_PUB","NPT4_PRIV","NPT4_PROG",
  "C150_4","C150_L4","C150_4_POOLED","C200_L4","C200_4"
)
print(data_dir)
```

**Import (base R):**

```{r}
# Update the path to your file as needed
inst_path <- file.path(data_dir, "Most-Recent-Cohorts-Institution.csv")

df <- read.csv(inst_path, stringsAsFactors = FALSE)
# Keep only columns that exist
df <- df[ , intersect(columns_to_load_institution, names(df))]

cat("Original shape:", nrow(df), "rows x", ncol(df), "cols\n")
head(df, 3)
```

------------------------------------------------------------------------

## 2) ROI Columns to Numeric & Missing Profile

```{r}
roi_cols <- c("COSTT4_A","MD_EARN_WNE_P10")
for (col in roi_cols) {
  # Coerce; non-numeric like 'PS' becomes NA
  df[[col]] <- suppressWarnings(as.numeric(df[[col]]))
}

cat("Missing values in ROI columns:\n")
print(colSums(is.na(df[roi_cols])))
```



------------------------------------------------------------------------

## 3) Profile Non-reporting (Missing COSTT4_A)

```{r}
df_missing_cost <- df[ is.na(df$COSTT4_A), ]

# CONTROL map (1/2/3)
control_map <- c("1"="Public","2"="Private Nonprofit","3"="Private For-Profit")
df_missing_cost$CONTROL_NAME <- control_map[ as.character(df_missing_cost$CONTROL) ]

cat(nrow(df_missing_cost), "institutions missing COSTT4_A\n")
ctrl_tb <- table(df_missing_cost$CONTROL_NAME)
ctrl_pct <- round(100 * prop.table(ctrl_tb), 2)
print(ctrl_tb); print(ctrl_pct)

# PREDDEG map
preddeg_map <- c(
  "0"="Not Classified",
  "1"="Predominantly Certificate",
  "2"="Predominantly Associate's",
  "3"="Predominantly Bachelor's",
  "4"="Exclusively Graduate"
)
df_missing_cost$PREDDEG_NAME <- preddeg_map[ as.character(df_missing_cost$PREDDEG) ]
table(df_missing_cost$PREDDEG_NAME)
```

Of the 3,182 institutions that are missing cost data, nearly two-thirds (65%) are Private For-Profit. 

These are mostly certificate-granting institutions, not traditional 4-year colleges.

------------------------------------------------------------------------

## 4) Cleaned Dataset (Drop NA in ROI)

```{r}
df_cleaned <- df[ complete.cases(df[roi_cols]), ]
cat("After drop-NA (ROI):", nrow(df_cleaned), "rows\n")
summary(df_cleaned[roi_cols])
```
After removing all rows with missing cost or earnings data, the final dataset is 3,075 institutions.

------------------------------------------------------------------------

## 5) Field-of-Study Import & Cleaning

```{r}
# Path placeholders; update as needed if available
fos_path <- file.path(data_dir, "Most-Recent-Cohorts-Field-of-Study.csv")

# If you don't have this file locally, guard with file.exists()
if (file.exists(fos_path)) {
  df_field <- read.csv(fos_path, stringsAsFactors = FALSE)
  
  # Keep relevant columns if present
  columns_to_load_field <- c(
    "EARN_GT_THRESHOLD_1YR","EARN_GT_THRESHOLD_5YR",
    "UNITID","CIPCODE","CREDLEV","CONTROL",
    "EARN_MDN_1YR","EARN_MDN_4YR"
  )
  df_field <- df_field[ , intersect(columns_to_load_field, names(df_field)) ]
  
  # Coerce and cap > 100 to NA
  make_num <- function(x) suppressWarnings(as.numeric(x))
  if ("EARN_GT_THRESHOLD_1YR" %in% names(df_field)) {
    df_field$EARN_GT_THRESHOLD_1YR <- make_num(df_field$EARN_GT_THRESHOLD_1YR)
    df_field$EARN_GT_THRESHOLD_1YR[df_field$EARN_GT_THRESHOLD_1YR > 100] <- NA
  }
  if ("EARN_GT_THRESHOLD_5YR" %in% names(df_field)) {
    df_field$EARN_GT_THRESHOLD_5YR <- make_num(df_field$EARN_GT_THRESHOLD_5YR)
    df_field$EARN_GT_THRESHOLD_5YR[df_field$EARN_GT_THRESHOLD_5YR > 100] <- NA
  }
  
  summary(df_field)
}
```


------------------------------------------------------------------------

## 6) Predominant Degree Share (Counts & Percent)

```{r}
df_cleaned$PREDDEG_NAME <- preddeg_map[ as.character(df_cleaned$PREDDEG) ]
roi_counts <- table(df_cleaned$PREDDEG_NAME)
roi_percent <- round(100 * prop.table(roi_counts), 2)
with_roi <- data.frame(Count = as.vector(roi_counts),
                       Percentage = as.vector(roi_percent),
                       row.names = names(roi_counts))
print(with_roi)
```

------------------------------------------------------------------------

## 7) Non-Traditional Institutions: Program Cost vs Earnings

```{r}
# Targets: Not Classified, Certificate, Associate's => codes 0,1,2
target_preddegs <- c(0,1,2)
# Ensure PREDDEG is numeric
df$PREDDEG <- suppressWarnings(as.numeric(df$PREDDEG))

df_missing_academic_cost <- df[ is.na(df$COSTT4_A), ]
df_target <- df_missing_academic_cost[ df_missing_academic_cost$PREDDEG %in% target_preddegs, ]

# Keep rows with both program cost & earnings
keep <- !is.na(df_target$COSTT4_P) & !is.na(df_target$MD_EARN_WNE_P10)
df_plot_ready <- df_target[ keep, ]

df_plot_ready$PREDDEG_NAME <- preddeg_map[ as.character(df_plot_ready$PREDDEG) ]

par(mar = c(2, 4.1, 2, 2))
# Base R scatter, color by PREDDEG_NAME
cols <- c("Not Classified"="black","Predominantly Certificate"="steelblue","Predominantly Associate's"="red")
plot(df_plot_ready$COSTT4_P, df_plot_ready$MD_EARN_WNE_P10,
     xlab="Average Annual Program Cost ($)", ylab="Median Earnings 10Y After Entry ($)",
     main="Program Cost vs Earnings (Non-Traditional Institutions)",
     col = cols[df_plot_ready$PREDDEG_NAME], pch=16, cex=0.7);
legend("topleft", legend=names(cols), col=cols, pch=16, bty="n")
```

------------------------------------------------------------------------

## 8) Institution Type & Scatter (Correct CONTROL Mapping)

```{r}
df_cleaned$CONTROL_NAME <- control_map[ as.character(df_cleaned$CONTROL) ]
inst_counts <- table(df_cleaned$CONTROL_NAME)
inst_pct <- round(100 * prop.table(inst_counts), 2)
data.frame(Count=as.vector(inst_counts), Percentage=as.vector(inst_pct),
           row.names = names(inst_counts))

# Cost vs Earnings colored by institution type
cols2 <- c("Public"="forestgreen","Private Nonprofit"="goldenrod","Private For-Profit"="firebrick")
plot(df_cleaned$COSTT4_A, df_cleaned$MD_EARN_WNE_P10,
     xlab="Cost of Attendance ($)", ylab="Median Earnings 10Y ($)",
     main="Cost vs Earnings by Institution Type",
     col=cols2[df_cleaned$CONTROL_NAME], pch=16, cex=0.6); 
legend("topleft", legend=names(cols2), col=cols2, pch=16, bty="n")
```

------------------------------------------------------------------------

## 9) Net Price, ROI, and Top-10

```{r}
# If public net price missing, use private net price
df$NPT4_PUB <- suppressWarnings(as.numeric(df$NPT4_PUB))
df$NPT4_PRIV <- suppressWarnings(as.numeric(df$NPT4_PRIV))

df$NET_PRICE <- ifelse(!is.na(df$NPT4_PUB), df$NPT4_PUB, df$NPT4_PRIV)
df_np <- df[ !is.na(df$NET_PRICE) & !is.na(df$MD_EARN_WNE_P10), ]
cat("Found", nrow(df_np), "institutions with Net Price & Earnings\n")

df_np$SCHOOL_TYPE <- control_map[ as.character(df_np$CONTROL) ]
df_np$ROI_RATIO <- df_np$MD_EARN_WNE_P10 / df_np$NET_PRICE

# Top-10 overall for earnings > 80k, highest ROI
top_overall <- df_np[ df_np$MD_EARN_WNE_P10 > 80000, c("INSTNM","ROI_RATIO","MD_EARN_WNE_P10","COSTT4_A","NET_PRICE") ]
top_overall <- top_overall[ order(-top_overall$ROI_RATIO), ]
head(top_overall, 10)
```

```{r}
# Cost vs Earnings colored by institution type
cols2 <- c("Public"="forestgreen","Private Nonprofit"="goldenrod","Private For-Profit"="firebrick")
plot(df_cleaned$NET_PRICE, df_cleaned$MD_EARN_WNE_P10,
     xlab="Net Price ($)", ylab="Median Earnings 10Y ($)",
     main="Net Price vs Earnings by Institution Type",
     col=cols2[df_cleaned$CONTROL_NAME], pch=16, cex=0.6)
grid(); legend("topleft", legend=names(cols2), col=cols2, pch=16, bty="n")

```

------------------------------------------------------------------------

## 10) Top-10 by Institution Type

```{r}
top_by_type <- function(ctrl_code) {
  tmp <- df_np[ df_np$CONTROL == ctrl_code & df_np$MD_EARN_WNE_P10 > 80000, 
                c("INSTNM","ROI_RATIO","MD_EARN_WNE_P10","COSTT4_A","NET_PRICE") ]
  tmp <- tmp[ order(-tmp$ROI_RATIO), ]
  head(tmp, 10)
}
top_public  <- top_by_type(1)
top_priv_np <- top_by_type(2)
top_priv_fp <- top_by_type(3)

top_public; top_priv_np; top_priv_fp
```

------------------------------------------------------------------------

## 11) Region Mapping, Scatter, and Means

```{r}
region_map <- c(
  "0"="U.S. Service Schools","1"="New England","2"="Mid East","3"="Great Lakes",
  "4"="Plains","5"="Southeast","6"="Southwest","7"="Rocky Mountains","8"="Far West","9"="Outlying Areas"
)
df_cleaned$NET_PRICE <- ifelse(!is.na(df_cleaned$NPT4_PUB), df_cleaned$NPT4_PUB, df_cleaned$NPT4_PRIV)
df_cleaned$REGION_NAME <- region_map[ as.character(df_cleaned$REGION) ]
table(df_cleaned$REGION_NAME)

# Scatter by region
reg_cols <- rainbow(length(unique(na.omit(df_cleaned$REGION_NAME))))
reg_lut <- setNames(reg_cols, sort(unique(na.omit(df_cleaned$REGION_NAME))))
plot(df_cleaned$NET_PRICE, df_cleaned$MD_EARN_WNE_P10,
     xlab="Net Price ($)", ylab="Median Earnings 10Y ($)",
     main="Net Price vs Earnings by Region",
     col=reg_lut[df_cleaned$REGION_NAME], pch=16, cex=0.6)
grid(); legend("topleft", legend=names(reg_lut), col=reg_lut, pch=16, cex=0.8, ncol=2, bty="n")

# Region means and diff
region_means <- aggregate(cbind(MD_EARN_WNE_P10, NET_PRICE) ~ REGION_NAME, df_cleaned, mean, na.rm=TRUE)
region_means$DIFF_ABS <- region_means$MD_EARN_WNE_P10 - region_means$NET_PRICE
region_means[ order(-region_means$DIFF_ABS), ]
```

------------------------------------------------------------------------

## 12) Region-Level Normalization & Trend

```{r}
# Normalize within region: (x - min) / (max - min)
norm_in_group <- function(x) (x - min(x, na.rm=TRUE)) / (max(x, na.rm=TRUE) - min(x, na.rm=TRUE))

df_cleaned$earn_norm <- ave(df_cleaned$MD_EARN_WNE_P10, df_cleaned$REGION_NAME, FUN=norm_in_group)
df_cleaned$net_norm  <- ave(df_cleaned$NET_PRICE,        df_cleaned$REGION_NAME, FUN=norm_in_group)
df_cleaned$net_diff_ratio <- df_cleaned$earn_norm - df_cleaned$net_norm

region_trends <- aggregate(net_diff_ratio ~ REGION_NAME, df_cleaned, mean, na.rm=TRUE)
region_trends <- region_trends[ order(-region_trends$net_diff_ratio), ]
region_trends

# Barplot
barplot(height = region_trends$net_diff_ratio, names.arg = region_trends$REGION_NAME,
        las=2, main="Relative Earnings - Cost Ratio by Region",
        ylab="Normalized (Earnings - Cost)", border=NA)
grid(nx=NA, ny=NULL)
```

------------------------------------------------------------------------

## 13) Locale Simplification & Plots

```{r}
simplify_locale <- function(locale_code) {
  if (is.na(locale_code)) return("Unknown")
  if (locale_code >= 11 && locale_code <= 13) return("City")
  if (locale_code >= 21 && locale_code <= 23) return("Suburb")
  if (locale_code >= 31 && locale_code <= 33) return("Town")
  if (locale_code >= 41 && locale_code <= 43) return("Rural")
  "Unknown"
}

df_cleaned$LOCALE_TYPE <- sapply(df_cleaned$LOCALE, simplify_locale)
table(df_cleaned$LOCALE_TYPE)

# Scatter by locale
loc_cols <- c("City"="dodgerblue3","Suburb"="darkorange2","Town"="purple","Rural"="darkgreen","Unknown"="gray60")
plot(df_cleaned$NET_PRICE, df_cleaned$MD_EARN_WNE_P10,
     xlab="Net Price ($)", ylab="Median Earnings 10Y ($)",
     main="Cost vs Earnings by Institutional Locale",
     col=loc_cols[df_cleaned$LOCALE_TYPE], pch=16, cex=0.6)
grid(); legend("topleft", legend=names(loc_cols), col=loc_cols, pch=16, bty="n")

# Locale summary means
loc_means <- aggregate(cbind(NET_PRICE, MD_EARN_WNE_P10) ~ LOCALE_TYPE, df_cleaned, mean, na.rm=TRUE)
loc_means$DIFF <- loc_means$MD_EARN_WNE_P10 - loc_means$NET_PRICE
loc_means
```

------------------------------------------------------------------------

## 14) UGDS Quartiles & Size Plot

```{r}
summary(df_cleaned$UGDS)

# Use exact cut points as in Python example
breaks <- c(0, 716, 1725, 4574, Inf)
labels <- c("Q1 (6-715)","Q2 (716-1725)","Q3 (1725-4574)","Q4 (>4574)")
df_cleaned$SCHOOL_SIZE <- cut(df_cleaned$UGDS, breaks=breaks, labels=labels, right=FALSE)

sz_cols <- setNames(rainbow(length(labels)), labels)
plot(df_cleaned$NET_PRICE, df_cleaned$MD_EARN_WNE_P10,
     xlab="Net Price ($)", ylab="Median Earnings 10Y ($)",
     main="Cost vs Earnings by UG Population",
     col=sz_cols[df_cleaned$SCHOOL_SIZE], pch=16, cex=0.6)
grid(); legend("topleft", legend=labels, col=sz_cols, pch=16, bty="n")
```

------------------------------------------------------------------------

## 15) ROI by Institution Type (Boxplot) & Medians

```{r}
# Ensure ROI present (df_np has NET_PRICE & MD_EARN_WNE_P10)
# Reuse df_np from earlier
if (!exists("df_np")) {
  df$NPT4_PUB <- suppressWarnings(as.numeric(df$NPT4_PUB))
  df$NPT4_PRIV <- suppressWarnings(as.numeric(df$NPT4_PRIV))
  df$NET_PRICE <- ifelse(!is.na(df$NPT4_PUB), df$NPT4_PUB, df$NPT4_PRIV)
  df_np <- df[ !is.na(df$NET_PRICE) & !is.na(df$MD_EARN_WNE_P10), ]
  df_np$SCHOOL_TYPE <- control_map[ as.character(df_np$CONTROL) ]
  df_np$ROI_RATIO <- df_np$MD_EARN_WNE_P10 / df_np$NET_PRICE
}

boxplot(ROI_RATIO ~ SCHOOL_TYPE, data=df_np,
        main="Return on Investment (ROI) by Institution Type",
        ylab="ROI Ratio", xlab="Institution Type",
        outline=FALSE)
abline(h=1, lty=2)
tapply(df_np$ROI_RATIO, df_np$SCHOOL_TYPE, median, na.rm=TRUE)
```

Calculating an ROI Based on Type of Institution, Public Institutions fare the best

------------------------------------------------------------------------

## 17) Completion Rate Cleaning & Overview

```{r}
df_with_cost <- df[!is.na(df$COSTT4_A), ]

completion_rates <- suppressWarnings(as.numeric(df_with_cost$C150_4_POOLED))

compl_clean <- completion_rates[!is.na(completion_rates)]

cat("Total institutions with cost data:", nrow(df_with_cost), "\n")
cat("Of those, institutions with valid completion data:", length(compl_clean), "\n")
cat("Missing rate within this subset (%):", round(100 * (1 - length(compl_clean) / nrow(df_with_cost)), 2), "\n")

summary(compl_clean)
```
Around 1/3rd of institutions with the 4 year cost of attendance don't have completion rates

------------------------------------------------------------------------

## 19) Completion & Earnings by Institution Type

```{r}
needed <- c("CONTROL","C150_4_POOLED","MD_EARN_WNE_P10")
if (all(needed %in% names(df))) {
  
  # Aggregate directly from the original 'df' data frame
  summary_by_type <- aggregate(
    cbind(C150_4_POOLED, MD_EARN_WNE_P10) ~ CONTROL, 
    data = df, 
    FUN = mean, 
    na.action = na.omit
  )
  
  # 2. Map the control codes to names for labeling.
  c(1, 2, 3)
  as.character( c(1, 2, 3) ) 
  c("Public", "Private Nonprofit", "Private For-Profit")
  summary_by_type$CONTROL_NAME <- c("Public", "Private Nonprofit", "Private For-Profit")
  
  print(summary_by_type)
  
  # Visualization
  plot_matrix <- t(as.matrix(
    data.frame(
      Completion = summary_by_type$C150_4_POOLED,
      Earnings_scaled = summary_by_type$MD_EARN_WNE_P10 / 100000
    )
  ))
  barplot(
    plot_matrix,
    beside = TRUE, 
    names.arg = summary_by_type$CONTROL_NAME,
    legend.text = c("Completion Rate", "Earnings (/100k)"),
    args.legend = list(x = "topleft", bty = "n"),
    col = c("skyblue", "orange"),
    main = "Avg Completion and Earnings by Institution Type",
    ylab = "Value (Rate or $/100k)"
  )
  grid(nx = NA, ny = NULL)
}
```

------------------------------------------------------------------------

## 20) Size vs Performance

```{r}
need2 <- c("UGDS","C150_4_POOLED","MD_EARN_WNE_P10")
if (all(need2 %in% names(df))) {
  size_df <- data.frame(
    UGDS=suppressWarnings(as.numeric(df$UGDS)),
    C150_4_POOLED=suppressWarnings(as.numeric(df$C150_4_POOLED)),
    MD_EARN_WNE_P10=suppressWarnings(as.numeric(df$MD_EARN_WNE_P10))
  )
  size_df <- size_df[ complete.cases(size_df), ]
  corr_size_completion <- cor(size_df$UGDS, size_df$C150_4_POOLED)
  corr_size_earnings <- cor(size_df$UGDS, size_df$MD_EARN_WNE_P10)
  cat(sprintf("Corr(UGDS, Completion): %.3f\n", corr_size_completion))
  cat(sprintf("Corr(UGDS, Earnings): %.3f\n", corr_size_earnings))
  
  par(mfrow=c(1,2))
  plot(size_df$UGDS, size_df$C150_4_POOLED, pch=16, col=rgb(0,0,0,0.4),
       xlab="UGDS", ylab="Completion Rate", main="Size vs Completion"); grid()
  plot(size_df$UGDS, size_df$MD_EARN_WNE_P10, pch=16, col=rgb(0,0,0,0.4),
       xlab="UGDS", ylab="Median Earnings 10Y ($)", main="Size vs Earnings"); grid()
  par(mfrow=c(1,1))
}
```

------------------------------------------------------------------------

## 21) Region Summary (Bars) & Success Index

```{r}
if (all(c("REGION","C150_4_POOLED","MD_EARN_WNE_P10") %in% names(df))) {
  region_df <- data.frame(
    REGION = suppressWarnings(as.numeric(df$REGION)),
    C150_4_POOLED = suppressWarnings(as.numeric(df$C150_4_POOLED)),
    MD_EARN_WNE_P10 = suppressWarnings(as.numeric(df$MD_EARN_WNE_P10))
  )
  region_df <- region_df[ complete.cases(region_df), ]
  region_df$REGION_NAME <- region_map[ as.character(region_df$REGION) ]
  
  summary_by_region <- aggregate(cbind(C150_4_POOLED, MD_EARN_WNE_P10) ~ REGION_NAME, region_df, mean)
  summary_by_region <- summary_by_region[ order(-summary_by_region$MD_EARN_WNE_P10), ]
  summary_by_region
  
  # Bars (earnings scaled)
  par(mar=c(10,4,3,1))
  mat <- t(as.matrix(cbind(summary_by_region$C150_4_POOLED, summary_by_region$MD_EARN_WNE_P10/100000)))
  barplot(mat, beside=TRUE, names.arg=summary_by_region$REGION_NAME, las=2,
          col=c("skyblue","orange"), legend.text=c("Completion","Earnings (/100k)"),
          main="Avg Completion and Earnings by Region")
  grid(nx=NA, ny=NULL)
  
  # Success Index (normalize columns 0–1 and average)
  rng_norm <- function(v) (v - min(v, na.rm=TRUE)) / (max(v, na.rm=TRUE) - min(v, na.rm=TRUE))
  summary_by_region$completion_norm <- rng_norm(summary_by_region$C150_4_POOLED)
  summary_by_region$earnings_norm   <- rng_norm(summary_by_region$MD_EARN_WNE_P10)
  summary_by_region$success_index   <- (summary_by_region$completion_norm + summary_by_region$earnings_norm)/2
  summary_by_region <- summary_by_region[ order(-summary_by_region$success_index), ]
}
```

------------------------------------------------------------------------

## 16) Field-of-Study (CIP) Area Summaries

```{r}

### This script calculates the median 4-year earnings per discipline,
### broken down by Associate's, Bachelor's, and Master's Degrees.
degree_column_name <- "CREDLEV_NAME"

# Replace privacy strings with NA
repl <- function(x) {
  x[x %in% c("PS", "PrivacySuppressed", "NULL", "None", "")] <- NA
  x
}
for (nm in names(df_field)) {
  df_field[[nm]] <- repl(df_field[[nm]])
}

num_cols <- intersect(c(
  "EARN_MDN_1YR", "EARN_MDN_4YR", "EARN_GT_THRESHOLD_1YR", "EARN_GT_THRESHOLD_5YR",
  "DEBT_MDN_SUPP", "NET_PRICE"
), names(df_field))
for (cname in num_cols) {
  df_field[[cname]] <- suppressWarnings(as.numeric(df_field[[cname]]))
}

# CIP2 and CIPAREA mapping
if ("CIPCODE" %in% names(df_field)) {
  df_field$CIP2 <- suppressWarnings(as.integer(as.numeric(df_field$CIPCODE) %/% 100))
  cip2_map <- c(
    "1" = "Agriculture & Related Sciences", "3" = "Natural Resources & Conservation",
    "4" = "Architecture & Related Services", "9" = "Communications & Journalism",
    "10" = "Communications Technologies", "11" = "Computer & Information Sciences",
    "13" = "Education", "14" = "Engineering", "15" = "Engineering Technologies",
    "19" = "Family & Consumer Sciences", "22" = "Legal Studies",
    "24" = "Liberal Arts & Humanities", "27" = "Mathematics & Statistics",
    "30" = "Multi/Interdisciplinary Studies", "38" = "Philosophy & Religious Studies",
    "40" = "Physical Sciences", "42" = "Psychology", "44" = "Public Administration & Social Services",
    "45" = "Social Sciences", "50" = "Visual & Performing Arts",
    "51" = "Health Professions & Related Programs",
    "52" = "Business, Management, & Marketing", "12" = "Personal & Culinary Services",
    "23" = "English Language & Literature", "31" = "Parks/Rec/Leisure & Kinesiology",
    "39" = "Theology & Religious Vocations", "46" = "Construction Trades",
    "48" = "Precision Production", "54" = "History"
  )
  df_field$CIPAREA <- cip2_map[as.character(df_field$CIP2)]
}


# These are the degree levels we want to include in our final table.
# IMPORTANT: The spelling must exactly match the values in your "CREDLEV_NAME" column.
desired_degrees <- c("Associate's", "Bachelor's", "Master's")
df_filtered <- df_field[df_field[[degree_column_name]] %in% desired_degrees, ]

# Aggregate median earnings by both discipline (CIPAREA) and degree level (CREDLEV_NAME)
summary_cip_degree <- aggregate(df_filtered["EARN_MDN_1YR"],
  by = list(
    CIPAREA = df_filtered$CIPAREA,
    Degree = df_filtered[[degree_column_name]]
  ),
  FUN = median,
  na.rm = TRUE
)

# Reshape the data from a long format to a wide format for easy comparison
earnings_by_degree <- reshape(summary_cip_degree,
  idvar = "CIPAREA",
  timevar = "Degree",
  direction = "wide"
)

earnings_by_degree_sorted <- earnings_by_degree[order(earnings_by_degree$`EARN_MDN_1YR.Bachelor's`, decreasing = TRUE), ]
earnings_by_degree_sorted
```
```{r}
# Master's earnings multiplier
earnings_by_degree_sorted$Masters_Multiplier <- earnings_by_degree_sorted$`EARN_MDN_1YR.Master's` / earnings_by_degree_sorted$`EARN_MDN_1YR.Bachelor's`

# raw earnings gain
earnings_by_degree_sorted$Masters_Raw_Gain <- earnings_by_degree_sorted$`EARN_MDN_1YR.Master's` - earnings_by_degree_sorted$`EARN_MDN_1YR.Bachelor's`

# Filter where multiplier or raw gain not calcualted
combined_data <- earnings_by_degree_sorted[
  !is.na(earnings_by_degree_sorted$Masters_Multiplier) & 
  is.finite(earnings_by_degree_sorted$Masters_Multiplier) &
  !is.na(earnings_by_degree_sorted$Masters_Raw_Gain), 
]

# descending order
combined_sorted <- combined_data[order(combined_data$Masters_Multiplier, decreasing = TRUE), ]


final_table <- data.frame(
  Discipline = combined_sorted$CIPAREA,
  Masters_Earnings_Multiplier = combined_sorted$Masters_Multiplier,
  Masters_Earnings_Gain = combined_sorted$Masters_Raw_Gain, # Corrected column name
  Bachelors_Median_Earnings = combined_sorted$`EARN_MDN_1YR.Bachelor's`,
  Masters_Median_Earnings = combined_sorted$`EARN_MDN_1YR.Master's`
)

print(final_table)
```

------------------------------------------------------------------------

## 22) Do Higher Degrees Lead to Higher Pay? ---


```{r top-visualize-1yrearnings-with-control}


cred_map <- c("1"="Ugrad Cert", "2"="Associate's", "3"="Bachelor's", "5"="Master's", "6"="Doctoral")
df_field$CREDLEV_NAME <- cred_map[as.character(df_field$CREDLEV)]
df_cred_compare <- df_field[df_field$CREDLEV %in% c(2, 3, 5), ]
earnings_by_cred <- aggregate(EARN_MDN_1YR ~ CIPAREA + CREDLEV_NAME, data = df_cred_compare, FUN = mean, na.rm = TRUE)

median_earnings_all <- aggregate(EARN_MDN_1YR ~ CREDLEV_NAME, data = df_cred_compare, FUN = median, na.rm = TRUE)

control_data <- c(
  "Associate's" = median_earnings_all$EARN_MDN_1YR[median_earnings_all$CREDLEV_NAME == "Associate's"],
  "Bachelor's"  = median_earnings_all$EARN_MDN_1YR[median_earnings_all$CREDLEV_NAME == "Bachelor's"],
  "Master's"    = median_earnings_all$EARN_MDN_1YR[median_earnings_all$CREDLEV_NAME == "Master's"]
)

example_cip <- c("Computer & Information Sciences", "Mathematics & Statistics")
plot_data_long <- earnings_by_cred[earnings_by_cred$CIPAREA %in% example_cip, ]

assoc_data <- plot_data_long[plot_data_long$CREDLEV_NAME == "Associate's", ]
bach_data  <- plot_data_long[plot_data_long$CREDLEV_NAME == "Bachelor's", ]
mast_data  <- plot_data_long[plot_data_long$CREDLEV_NAME == "Master's", ]

assoc_data <- assoc_data[order(assoc_data$CIPAREA), ]
bach_data  <- bach_data[order(bach_data$CIPAREA), ]
mast_data  <- mast_data[order(mast_data$CIPAREA), ]

plot_matrix <- cbind(
  "Associate's" = assoc_data$EARN_MDN_1YR,
  "Bachelor's"  = bach_data$EARN_MDN_1YR,
  "Master's"    = mast_data$EARN_MDN_1YR
)
rownames(plot_matrix) <- assoc_data$CIPAREA


plot_matrix_with_control <- rbind("All Disciplines" = control_data, plot_matrix)

wrapped_labels <- sapply(rownames(plot_matrix_with_control), function(x) {
  paste(strwrap(x, width = 25), collapse = "\n")
})

# sisable scientific notation
options(scipen = 999)

par(mar = c(5, 10, 4, 6))
max_val <- max(plot_matrix_with_control, na.rm = TRUE) * 1.1

barplot(
  height = t(plot_matrix_with_control),
  beside = TRUE,
  horiz = TRUE,
  main = "Where Does Data Science Fit In?",
  xlab = "Median Earnings 1 Year Post-Completion (2018-2021)",
  names.arg = wrapped_labels, 
  las = 1,
  cex.names = 0.7,
  col = c("skyblue", "steelblue", "darkblue"),
  xlim = c(0, max_val)
)

# uncrop the plot
par(xpd=TRUE)

legend(
  "topright", 
  inset = c(-0.2, 0),
  legend = c("Associate's", "Bachelor's", "Master's"),
  fill = c("skyblue", "steelblue", "darkblue"),
  title = "Degree Level",
  bty = "n",
  cex = 0.8
)
```

```{r top-visualize-4yrearnings-with-control}

earnings_by_cred <- aggregate(EARN_MDN_4YR ~ CIPAREA + CREDLEV_NAME, data = df_cred_compare, FUN = mean, na.rm = TRUE)

median_earnings_all <- aggregate(EARN_MDN_4YR ~ CREDLEV_NAME, data = df_cred_compare, FUN = median, na.rm = TRUE)

control_data <- c(
  "Associate's" = median_earnings_all$EARN_MDN_4YR[median_earnings_all$CREDLEV_NAME == "Associate's"],
  "Bachelor's"  = median_earnings_all$EARN_MDN_4YR[median_earnings_all$CREDLEV_NAME == "Bachelor's"],
  "Master's"    = median_earnings_all$EARN_MDN_4YR[median_earnings_all$CREDLEV_NAME == "Master's"]
)

example_cip <- c("Computer & Information Sciences", "Mathematics & Statistics")
plot_data_long <- earnings_by_cred[earnings_by_cred$CIPAREA %in% example_cip, ]

assoc_data <- plot_data_long[plot_data_long$CREDLEV_NAME == "Associate's", ]
bach_data  <- plot_data_long[plot_data_long$CREDLEV_NAME == "Bachelor's", ]
mast_data  <- plot_data_long[plot_data_long$CREDLEV_NAME == "Master's", ]

assoc_data <- assoc_data[order(assoc_data$CIPAREA), ]
bach_data  <- bach_data[order(bach_data$CIPAREA), ]
mast_data  <- mast_data[order(mast_data$CIPAREA), ]

plot_matrix <- cbind(
  "Associate's" = assoc_data$EARN_MDN_4YR,
  "Bachelor's"  = bach_data$EARN_MDN_4YR,
  "Master's"    = mast_data$EARN_MDN_4YR
)
rownames(plot_matrix) <- assoc_data$CIPAREA


plot_matrix_with_control <- rbind("All Disciplines" = control_data, plot_matrix)

wrapped_labels <- sapply(rownames(plot_matrix_with_control), function(x) {
  paste(strwrap(x, width = 25), collapse = "\n")
})

# sisable scientific notation
options(scipen = 999)

par(mar = c(5, 10, 4, 6))
# hacky but w/e its late
max_val <- max(plot_matrix_with_control, na.rm = TRUE) * 1.15

barplot(
  height = t(plot_matrix_with_control),
  beside = TRUE,
  horiz = TRUE,
  main = "Where Does Data Science Fit In? (2)",
  xlab = "Median Earnings 4 Year Post-Completion (2018-2021)",
  names.arg = wrapped_labels, 
  las = 1,
  cex.names = 0.7,
  col = c("skyblue", "steelblue", "darkblue"),
  xlim = c(0, max_val)
)

# uncrop the plot
par(xpd=TRUE)

legend(
  "topright", 
  inset = c(-0.2, 0),
  legend = c("Associate's", "Bachelor's", "Master's"),
  fill = c("skyblue", "steelblue", "darkblue"),
  title = "Degree Level",
  bty = "n",
  cex = 0.8
)
```



